---
output:
  ioslides_presentation:
    widescreen: true
    self_contained: true
---

```{r include = FALSE}
# Set options for knitr at the beginning of the document
knitr::opts_chunk$set(cache = TRUE)

# Read in configuration variables, but remove and hard code the worksheet
library(yaml)
config <- yaml.load_file("_config.yml")
config$handouts <- gsub('.', paste0('-', config$lesson, '.'),
                        config$handouts, fixed = TRUE)
```

# Database to Document with RMarkdown

Instructor: `r config$instructor`

## The End of the Pipeline

The pipeline under construction begins with data, moves through processing, analysis and visualization, and typically ends by producing content for a final report. 

RMarkdown offers the tantalizing opportunity to merge content and produce that final report as well.

## Objectives

1. Use the R and RMarkdown languages to script an entire pipeline in one file.
1. Connect the input of the pipeline to a database.
1. Publish the output of the pipeline on GitHub Pages.

## RMarkdown

RMarkdown, like R itself, is both a language and an interpreter.

**Language:** a set of special characters and structural rules that accompany text as instructions to the interpreter

**Interpreter:** software that converts the text and instructions into a formatted document of specified type

## RMarkdown + `knitr`

RMarkdown calls the `knitr` package to excute everything fenced off as R script within an RMarkdown document, and then selectively includes results in the formatted document.

## Seeing is believing

Open the handout `r config$handouts`, which is the RMarkdown source for this presentation. 

```{r}
vals <- c(4, 5, 6)
data <- data.frame(counts = vals)
data
```

The print out of `data` is not in the source; it was "knit" into the final document.

## Worksheet

The RMarkdown source for this lesson, with portions replaced by `...` that are for you to repair, is your worksheet. The goal of each exercise is to write the source for the generated slide.

## Exercise 1

To compare the output of your worksheet to this slide, press the "Knit" button in RStudio.

## RMarkdown Syntax

A RMarkdown document must begin with header content (or nothing) between a pair of lines containing only `---`. The header specifies the type of document to create and other configurations:

    ---
    output:
      html_document:
        self-contained: true
    ---

## RMarkdown Syntax

A line beginning with `#` symbols denotes headers of varying size

    # The Biggest Heading
    ## The Second Biggest Heading
    ### The Third Biggest Heading

Block quotes, like this syntax example, are produced by indenting text 4 spaces. Text within a block quote is displayed verbatim (as above). Without indenting, formatting happens (as follows).

# The Biggest Heading
## The Second Biggest Heading
### The Third Biggest Heading

## RMarkdown Syntax

Separate text with pipes ("|") to indicate columns of a table and hyphens ("-") to mark the beginning of a table or to separate the header.

    id | treatment
    ---|----------
    1  | control
    2  | exclosure

id | treatment
---|----------
1  | control
2  | exclosure
    
## RMarkdown Syntax

Embedding text between pairs of special characters induces some simple formatting. For example, use \* for *italics* and \*\* for **bold**.

## Exercise 2

charcter | format
---------|----------
*        | *italics*
**       | **bold**
^        | super^script^
~~       | ~~strikethrough~~

## RMarkdown Syntax

Many more elements of markdown await discovery:

1. lists
1. [links](http://lmgtfy.com/?q=markdown+cheat+sheet)
1. ... but let's move on to embedding R code!

## Code chunks

A R "code chunk"" is fenced with ` ```{r} ` above and ` ``` ` below the code.

> \`\`\`{r}  
> vals <- c(4, 5, 6)  
> \`\`\`

That's three backtick characters, found next to the "1" on QWERTY keyboards.

## Code chunks

Options, such as `include = FALSE` or `eval = FALSE`, specified at each chunk will override the defaults. They are place after the `r` and optional chunk name at the top of each code chunk.

For example:

> \`\`\`{r assign_vals, echo = FALSE}  
> vals <- c(4, 5, 6)  
> \`\`\`

will (by default) evaluate the content but not show the code in the document.

## Code chunks

A value defined in one code chunk, i.e.

```{r}
vals <- c(4, 5, 6)
```

is accessible in another:

```{r}
vals + 1
```

## Exercise 3: Prelude

Change the settings at the beginning of your worksheet to allow R code evaluation:

> \`\`\`{r include = FALSE}
> # Set options for knitr at the beginning of the document
> knitr::opts_chunk$set(cache = TRUE)
>
> # Delete the following setting upon reaching Exercise 3
> knitr::opts_chunk$set(eval = FALSE)
> \`\`\`

## Exercise 3

```{r include = FALSE}
library(dplyr)
library(RPostgreSQL)
con <- dbConnect(PostgreSQL(), dbname = 'portal', host = 'localhost')
animals <- tbl(con, 'animals')
```

```{r}
animals
```

## Construct the Pipeline

If your entire pipeline can be scripted in R, then you could simply paste a long script into a single code chunk before moving on to generating content (e.g. tables or plots) for the final document.

It's better to modularize your script: break it into separate chunks that perform discrete tasks and let RMarkdown cache intermediate results.

```{r wrangle, include = FALSE}
# get taxonomic data
species <- tbl(con, 'species')

# group animals captured in 1990 by species_id
# and summarize with total abundance and average weight
# ignoring non-rodents
rodents <- filter(animals, year == 1990) %>%
  inner_join(species, c('species_id' = 'id')) %>%
  filter(taxa == 'Rodent') %>%
  group_by(species_id) %>%
  summarize(count = n(), weight = mean(weight)) %>%
  collect()

# disconnect from the PostgreSQL server
dbDisconnect(con)
```

```{r model, dependson = c('wrangle'), include = FALSE}
# estimate coefficients for a simple, linear regression
model <- lm(count ~ weight, data = rodents)
```

```{r viz, dependson = c('model'), include = FALSE}
library(ggplot2)

figA <- rodents %>%
  mutate(species_id = reorder(species_id, weight)) %>%
  ggplot(aes(x = species_id, y = weight)) +
  geom_bar(stat = 'identity')

figB <- rodents %>%
  mutate(species_id = reorder(species_id, weight)) %>%
  ggplot(aes(x = species_id, y = count)) +
  geom_bar(stat = 'identity')
```

## My Portal Study

Finally, write up your analysis of the data. Include nicely formatted tables of fitted models.

```{r include = FALSE}
library(stargazer)
```

```{r results = 'asis', dependson = c('model'), echo = FALSE}
stargazer(model, type = 'html')
```

## My Portal Study

Adjust your figures to fit nicely in the output document, but don't expect too much.

```{r include = FALSE}
library(gridExtra)
```

```{r dependson = c('viz'), echo = FALSE, warning = FALSE}
grid.arrange(figA, figB, ncol = 2)
```

## Publish to GitHub Pages

Your handouts repository on GitHub includes a free web hosting service known as GitHub Pages. Publish your worksheet there with the following steps.

- Copy the resulting HTML file to 'docs/index.html'
- Notice the essential (but empty) 'docs/_config.yml' file
- Commit & Push
- Enable GitHub Pages: Settings > master/docs folder

## Solutions

## Solution 1

To compare the output of your worksheet to this slide, press the "Knit" button in RStudio.

## Solution 2

charcter | format
---------|----------
*        | *italics*
**       | **bold**
^        | super^script^
~~       | ~~strikethrough~~

## Solution 3

```{r include = FALSE}
library(dplyr)
library(RPostgreSQL)
con <- dbConnect(PostgreSQL(), dbname = 'portal', host = 'localhost')
animals <- tbl(con, 'animals')
```

```{r}
animals
```

```{r include = FALSE}
# this is just cleanup, not part of the solution
dbDisconnect(con)
```